---
title: "Marketplace Revenue Risk Analysis"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

# Marketplace Revenue Risk Analysis

## Setup

This section loads required libraries and connects to the database.

```{r setup, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
library(ggplot2)
library(dplyr)

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "leowan34",
  host = "localhost",
  port = 5332,
  user = "leowan34",
  password = Sys.getenv("PG_PASSWORD")
)
```
## Data Loading

Loading the datasets from the database for analysis.
```{r data-loading}
monthly <- dbReadTable(con, "monthly_revenue")
customer <- dbReadTable(con, "customer_revenue_decile")
payment <- dbReadTable(con, "order_payment_structure")
```



## Revenue Stability

This section evaluates the volatility of monthly revenue and short-term growth dynamics.
```{r}
sd_rev <- sd(monthly$monthly_revenue, na.rm = TRUE)
mean_rev <- mean(monthly$monthly_revenue, na.rm = TRUE)
cv_rev <- sd_rev / mean_rev

cv_rev
```
The coefficient of variation indicates moderate revenue volatility, showing that revenue is growing but not fully stable.

```{r}
monthly <- monthly %>%
  arrange(month) %>%
  mutate(
    mom_growth = (monthly_revenue - lag(monthly_revenue)) /
      lag(monthly_revenue)
  )
monthly_growth <- monthly %>%
  filter(!is.na(mom_growth))

sd(monthly$mom_growth, na.rm = TRUE)
```
The month-on-month growth rate shows significant fluctuations, showing a unstable performance in the short term.

```{r warning=FALSE}
ggplot(monthly, aes(x = month, y = monthly_revenue)) +
  geom_line(linewidth = 1.2, color = "#2C3E50") +
  geom_point(size = 2, color = "#1ABC9C") +
  labs(
    title = "Monthly Revenue Trend",
    x = "Month",
    y = "Revenue"
  ) +
  theme_minimal()
```

The trend shows a overall revenue growth, with a noticeable spike and some volatility near the later period.

```{r}
ggplot(monthly_growth, aes(x = month, y = mom_growth)) +
  geom_col(fill = "#3498DB") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Month-over-Month Revenue Growth",
    x = "Month",
    y = "MoM Growth Rate"
  ) +
  theme_minimal()
```

MoM growth is highly uneven, with strong positive jumps and several negative months.

## Geographic Concentration

This section examines the degree of geographic concentration in revenue and the potential risk of over-reliance on specific regions.
```{r}
state <- dbReadTable(con, "state_revenue")

state <- state %>%
  arrange(desc(total_revenue)) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(state, aes(x = reorder(customer_state, revenue_share),
                  y = revenue_share)) +
  geom_col(fill = "#3498DB") +
  coord_flip() +
  labs(
    title = "Revenue Share by State",
    x = "State",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is highly concentrated in a few key states, with São Paulo (SP) alone contributing approximately 37% of total revenue, followed by Rio de Janeiro (RJ) and Minas Gerais (MG) at around 13% and 12%, respectively.

```{r}
hhi <- sum(state$revenue_share^2)
hhi
```
The HHI value is approximately 0.18, indicating a moderate level of geographic concentration. This suggests that revenue is not fully diversified and still relies on a few key regions.


## Customer Concentration

This section measures how revenue is distributed across customers and whether a small share of customers contributes a disproportionate share of revenue.

```{r}
customer <- dbReadTable(con, "customer_revenue_decile")
```

```{r}
decile_summary <- customer %>%
  group_by(decile) %>%
  summarise(
    customer_count = n(),
    total_revenue = sum(total_revenue),
    .groups = "drop"
  ) %>%
  arrange(decile) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(decile_summary, aes(x = factor(decile), y = revenue_share)) +
  geom_col(fill = "#9B59B6") +
  labs(
    title = "Revenue Share by Customer Decile",
    x = "Customer Decile",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is concentrated in the top customer deciles. The first decile alone contributes about 38% of total revenue, while the top three deciles together contribute roughly 64%, indicating a strong dependence on a relatively small group of high-value customers.

```{r}
customer_sort <- customer %>%
  arrange(total_revenue) %>%
  mutate(
    cum_customer = row_number() / n(),
    cum_revenue = cumsum(total_revenue) / sum(total_revenue)
  )
```

```{r}
ggplot(customer_sort, aes(x = cum_customer, y = cum_revenue)) +
  geom_line(color = "#E67E22", linewidth = 1.2) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "Lorenz Curve of Customer Revenue Distribution",
    x = "Cumulative Share of Customers",
    y = "Cumulative Share of Revenue"
  ) +
  theme_minimal()
```

The Lorenz curve bends far away from the equality line, showing clear inequality in revenue contribution across customers.

```{r}
library(ineq)
gini <- ineq(customer$total_revenue, type = "Gini")
gini
```

The Gini coefficient is approximately 0.475, which suggests a moderate-to-high level of customer concentration.


## Liquidity Risk

This section evaluates whether payment structure could create cash flow timing risk.

```{r}
payment <- dbReadTable(con, "order_payment_structure")
```

```{r}
payment_summary <- payment %>%
  group_by(payment_type) %>%
  summarise(
    order_count = n(),
    total_revenue = sum(order_total),
    .groups = "drop"
  ) %>%
  mutate(
    order_share = order_count / sum(order_count),
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(payment_summary, aes(x = payment_type, y = revenue_share)) +
  geom_col(fill = "#1ABC9C") +
  labs(
    title = "Revenue Share by Payment Type",
    x = "Payment Type",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Installment payments account for a larger share of revenue than one-time payments, showing that cash flow may be collected over a longer period rather than immediately.

```{r}
weighted_installment <- sum(payment$max_installments * payment$order_total) /
  sum(payment$order_total)

weighted_installment
```

The weighted average installment count is approximately 4.12, which implys that installment-based purchases are actually common and may increase liquidity timing risk.

## Key Findings 
- Revenue shows moderate volatility, indicating instability in short-term performance. 
- Revenue is geographically concentrated, with São Paulo contributing approximately 37% of total revenue. 
- Customer concentration is significant, with the top decile contributing about 38% of total revenue and a Gini coefficient of around 0.47. 
- Payment structure indicates liquidity risk, with installment payments dominating revenue and an average installment count of approximately 4.1. 

Overall, the business faces moderate financial risk due to concentration and cash flow timing factors.
