---
title: "Marketplace Revenue Risk & Financial Stability Analysis"
author: "Liyuan Wan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

## Setup

This section loads required libraries and connects to the database.

```{r setup, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
library(ggplot2)
library(dplyr)
library(ineq)

con <- dbConnect(
  RPostgres::Postgres(),
  dbname   = Sys.getenv("OLIST_DB_NAME", "leowan34"),
  host     = Sys.getenv("OLIST_DB_HOST", "localhost"),
  port     = as.integer(Sys.getenv("OLIST_DB_PORT", "5332")),
  user     = Sys.getenv("OLIST_DB_USER", "leowan34"),
  password = Sys.getenv("PG_PASSWORD")
)
```

## Data Loading

Loading the datasets from the database for analysis.

```{r data-loading}
monthly  <- dbReadTable(con, "monthly_revenue")
customer <- dbReadTable(con, "customer_revenue_decile")
payment  <- dbReadTable(con, "order_payment_structure")
state    <- dbReadTable(con, "state_revenue")
```

---

## Revenue Stability

This section evaluates the volatility of monthly revenue and short-term growth dynamics to assess how predictable cash flows are over time.

```{r}
sd_rev   <- sd(monthly$monthly_revenue, na.rm = TRUE)
mean_rev <- mean(monthly$monthly_revenue, na.rm = TRUE)
cv_rev   <- sd_rev / mean_rev

cv_rev
```

The coefficient of variation (CV = `r round(cv_rev, 2)`) indicates moderate revenue volatility. While the overall trend is upward, revenue is not yet stable enough for reliable short-term forecasting — a meaningful consideration for cash flow planning and budgeting.

```{r}
monthly <- monthly %>%
  arrange(month) %>%
  mutate(
    mom_growth = (monthly_revenue - lag(monthly_revenue)) / lag(monthly_revenue)
  )

monthly_growth <- monthly %>%
  filter(!is.na(mom_growth))

sd(monthly$mom_growth, na.rm = TRUE)
```

Month-over-month growth shows high standard deviation, reflecting an unstable short-term performance pattern. Large swings between consecutive months suggest the business has not yet achieved consistent growth momentum.

```{r warning=FALSE}
ggplot(monthly, aes(x = month, y = monthly_revenue)) +
  geom_line(linewidth = 1.2, color = "#2C3E50") +
  geom_point(size = 2, color = "#1ABC9C") +
  labs(
    title = "Monthly Revenue Trend (2017–2018)",
    x = "Month",
    y = "Revenue (BRL)"
  ) +
  theme_minimal()
```

Revenue grew steadily through 2017, with a sharp spike in late 2017 followed by elevated but volatile performance through 2018. The spike warrants further investigation — it may reflect a seasonal event or promotional campaign rather than organic growth.

```{r}
ggplot(monthly_growth, aes(x = month, y = mom_growth)) +
  geom_col(fill = "#3498DB") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Month-over-Month Revenue Growth Rate",
    x = "Month",
    y = "MoM Growth Rate"
  ) +
  theme_minimal()
```

MoM growth is highly uneven — several months of strong positive growth are offset by multiple months of decline. This pattern is consistent with a marketplace still in an early growth stage, subject to demand shocks and inconsistent order volume.

---

## Geographic Concentration

This section examines the degree of geographic concentration in revenue and the risk of over-reliance on specific regions.

```{r}
state <- state %>%
  arrange(desc(total_revenue)) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(state, aes(x = reorder(customer_state, revenue_share),
                  y = revenue_share)) +
  geom_col(fill = "#3498DB") +
  coord_flip() +
  labs(
    title = "Revenue Share by State",
    x = "State",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is highly concentrated in a small number of states. São Paulo (SP) alone accounts for approximately 37% of total revenue, followed by Rio de Janeiro (RJ) at ~13% and Minas Gerais (MG) at ~12%. The remaining 24 states collectively contribute less than 40%, indicating that the marketplace's financial performance is closely tied to conditions in Southeast Brazil.

```{r}
hhi <- sum(state$revenue_share^2)
hhi
```

The Herfindahl-Hirschman Index (HHI = `r round(hhi, 3)`) reflects a moderate level of geographic concentration. HHI is adapted here from its antitrust convention (firm market shares) to state-level revenue shares; the U.S. DOJ moderately-concentrated band of 0.15–0.25 is used as a reference range. While the market is not monopolistically concentrated, the heavy dependence on SP introduces regional exposure — any economic disruption, logistics issue, or regulatory change in São Paulo could materially impact overall revenue.

---

## Customer Concentration

This section measures how revenue is distributed across customers and whether a small share of customers contributes a disproportionate share of revenue.

```{r}
decile_summary <- customer %>%
  group_by(decile) %>%
  summarise(
    customer_count = n(),
    total_revenue  = sum(total_revenue),
    .groups = "drop"
  ) %>%
  arrange(decile) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(decile_summary, aes(x = factor(decile), y = revenue_share)) +
  geom_col(fill = "#9B59B6") +
  labs(
    title = "Revenue Share by Customer Decile",
    x    = "Customer Decile (1 = Highest Spenders)",
    y    = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is heavily concentrated among the highest-value customers. The top decile alone contributes approximately 38% of total revenue, and the top three deciles together account for roughly 64%. This steep drop-off across deciles signals a high degree of customer dependency — losing a small number of high-value customers could have an outsized impact on total revenue.

```{r}
customer_sort <- customer %>%
  arrange(total_revenue) %>%
  mutate(
    cum_customer = row_number() / n(),
    cum_revenue  = cumsum(total_revenue) / sum(total_revenue)
  )
```

```{r}
ggplot(customer_sort, aes(x = cum_customer, y = cum_revenue)) +
  geom_line(color = "#E67E22", linewidth = 1.2) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "Lorenz Curve of Customer Revenue Distribution",
    x = "Cumulative Share of Customers",
    y = "Cumulative Share of Revenue"
  ) +
  theme_minimal()
```

The Lorenz curve deviates substantially from the line of perfect equality, confirming a highly unequal distribution of revenue across the customer base.

```{r}
gini <- ineq(customer$total_revenue, type = "Gini")
gini
```

The Gini coefficient of `r round(gini, 3)` indicates a moderate-to-high level of customer revenue inequality. This is consistent with typical marketplace dynamics where a small cohort of repeat, high-value buyers drives the majority of revenue — but also represents a concentration risk if those customers churn.

---

## Cash Flow Timing

This section evaluates whether the payment structure creates a measurable gap between revenue recognition and cash collection. Note that "liquidity risk" in the strict finance sense means inability to convert assets to cash; what this section measures is cash-collection timing driven by installment behavior, which sits closer to working-capital cycle analysis than to liquidity risk per se.

```{r}
payment_summary <- payment %>%
  group_by(payment_type) %>%
  summarise(
    order_count   = n(),
    total_revenue = sum(order_total),
    .groups = "drop"
  ) %>%
  mutate(
    order_share   = order_count / sum(order_count),
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(payment_summary, aes(x = payment_type, y = revenue_share)) +
  geom_col(fill = "#1ABC9C") +
  labs(
    title = "Revenue Share by Payment Type",
    x = "Payment Type",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Installment payments account for a significantly larger share of revenue than one-time payments. This means cash is collected incrementally over time rather than upfront, creating a structural gap between when revenue is recognized and when it is actually received.

```{r}
weighted_installment <- sum(payment$max_installments * payment$order_total) /
  sum(payment$order_total)

weighted_installment
```

The revenue-weighted average installment count is approximately `r round(weighted_installment, 2)`. Brazilian e-commerce operates within the "12x sem juros" (12-installment interest-free) cultural baseline, so 4.12 is modest by local benchmarks; the dominant cash-timing signal in this dataset is the share of revenue routed through installments (~64%), not the per-order installment depth. The structural implication is that a meaningful portion of recognized revenue is collected weeks to months after the transaction, lengthening the working-capital cycle.

---

## Final Summary

This analysis quantified four dimensions of financial risk on the Olist Brazilian e-commerce dataset (99,442 orders, 2016–2018):

- **Revenue Stability**: Moderate volatility (CV = 0.42) with uneven short-term growth
- **Geographic Concentration**: SP drives 37% of revenue (HHI = 0.18 — adapted from antitrust convention to state-level shares), creating regional exposure
- **Customer Concentration**: Top 10% of customers account for 38% of revenue (Gini = 0.475)
- **Cash Flow Timing**: Installment-heavy payment structure (~64% of revenue, avg. 4.12 installments) lengthens the working-capital cycle
- **Forecast**: Short-term revenue is expected to remain at elevated levels, with meaningful uncertainty reflecting historical volatility

Together, these findings describe a business in a growth phase with concentration and cash-timing patterns that warrant active monitoring as it scales.

---

## Patterns Worth Further Investigation

This section reframes the analytical findings into observable patterns and the business questions they raise. Recommendations on what to *do* about each pattern sit with PMs and business owners; the role of this analysis is to surface the patterns and quantify them.

- **Revenue volatility (CV = 0.42)**: The largest negative MoM swings occur in mid-2017 (June–July) and around the December 2017 → January 2018 transition. Mid-year aligns with the post-Carnival demand softening characteristic of Brazilian retail seasonality; the late-year drop aligns with post-Black-Friday / post-holiday fatigue. The pattern suggests revenue volatility has a partially predictable seasonal component, which is observable in the monthly_revenue volatility_signal field. The question worth carrying into a planning conversation: how much of the CV is structurally seasonal versus driven by month-specific shocks?

- **Geographic concentration (HHI = 0.18, SP = 37%)**: SP's 37% share creates regional exposure; logistics, economic, or regulatory shocks in São Paulo would be transmitted disproportionately to the full P&L. MG (~12%) and PR (~5%) appear under-represented relative to their share of national GDP (MG ≈ 9% of Brazilian GDP, PR ≈ 6%), making them the two most observable diversification gaps in the data. The pattern raises a question rather than prescribing an answer: is the under-representation a demand-side gap (untapped buyers) or a supply-side gap (insufficient sellers)? The dataset itself cannot distinguish the two.

- **Customer concentration (Gini = 0.475)**: The Lorenz curve's deviation from equality is driven primarily by the top two deciles, which together hold roughly 52% of revenue. The top decile alone holds 38%; losing the top 1% of customers would shift the revenue base by a measurable margin. The pattern surfaces a retention-sensitivity question — at what churn rate among the top decile would the overall revenue base materially shift? — but the dataset does not contain churn observations, so this remains a hypothesis worth testing with longitudinal data the marketplace owns.

- **Cash flow timing (Avg installments = 4.12, installment revenue share ≈ 64%)**: The per-order installment depth (4.12) is modest within Brazil's "12x sem juros" baseline; the dominant cash-timing signal is the share of revenue routed through installments (~64%), which lengthens the working-capital cycle relative to upfront-paid revenue. The observation worth raising: cash collection timing is structurally driven by payment-channel mix, so the installment-share trend is the more informative signal for working-capital sizing than the headcount of installment orders.
