---
title: "Marketplace Revenue Risk & Financial Stability Analysis"
author: "Liyuan Wan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

## Setup

This section loads required libraries and connects to the database.

```{r setup, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
library(ggplot2)
library(dplyr)
library(ineq)

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "leowan34",
  host = "localhost",
  port = 5332,
  user = "leowan34",
  password = Sys.getenv("PG_PASSWORD")
)
```

## Data Loading

Loading the datasets from the database for analysis.

```{r data-loading}
monthly  <- dbReadTable(con, "monthly_revenue")
customer <- dbReadTable(con, "customer_revenue_decile")
payment  <- dbReadTable(con, "order_payment_structure")
state    <- dbReadTable(con, "state_revenue")
```

---

## Revenue Stability

This section evaluates the volatility of monthly revenue and short-term growth dynamics to assess how predictable cash flows are over time.

```{r}
sd_rev   <- sd(monthly$monthly_revenue, na.rm = TRUE)
mean_rev <- mean(monthly$monthly_revenue, na.rm = TRUE)
cv_rev   <- sd_rev / mean_rev

cv_rev
```

The coefficient of variation (CV = `r round(cv_rev, 2)`) indicates moderate revenue volatility. While the overall trend is upward, revenue is not yet stable enough for reliable short-term forecasting — a meaningful consideration for cash flow planning and budgeting.

```{r}
monthly <- monthly %>%
  arrange(month) %>%
  mutate(
    mom_growth = (monthly_revenue - lag(monthly_revenue)) / lag(monthly_revenue)
  )

monthly_growth <- monthly %>%
  filter(!is.na(mom_growth))

sd(monthly$mom_growth, na.rm = TRUE)
```

Month-over-month growth shows high standard deviation, reflecting an unstable short-term performance pattern. Large swings between consecutive months suggest the business has not yet achieved consistent growth momentum.

```{r warning=FALSE}
ggplot(monthly, aes(x = month, y = monthly_revenue)) +
  geom_line(linewidth = 1.2, color = "#2C3E50") +
  geom_point(size = 2, color = "#1ABC9C") +
  labs(
    title = "Monthly Revenue Trend (2017–2018)",
    x = "Month",
    y = "Revenue (BRL)"
  ) +
  theme_minimal()
```

Revenue grew steadily through 2017, with a sharp spike in late 2017 followed by elevated but volatile performance through 2018. The spike warrants further investigation — it may reflect a seasonal event or promotional campaign rather than organic growth.

```{r}
ggplot(monthly_growth, aes(x = month, y = mom_growth)) +
  geom_col(fill = "#3498DB") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Month-over-Month Revenue Growth Rate",
    x = "Month",
    y = "MoM Growth Rate"
  ) +
  theme_minimal()
```

MoM growth is highly uneven — several months of strong positive growth are offset by multiple months of decline. This pattern is consistent with a marketplace still in an early growth stage, subject to demand shocks and inconsistent order volume.

---

## Geographic Concentration

This section examines the degree of geographic concentration in revenue and the risk of over-reliance on specific regions.

```{r}
state <- state %>%
  arrange(desc(total_revenue)) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(state, aes(x = reorder(customer_state, revenue_share),
                  y = revenue_share)) +
  geom_col(fill = "#3498DB") +
  coord_flip() +
  labs(
    title = "Revenue Share by State",
    x = "State",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is highly concentrated in a small number of states. São Paulo (SP) alone accounts for approximately 37% of total revenue, followed by Rio de Janeiro (RJ) at ~13% and Minas Gerais (MG) at ~12%. The remaining 24 states collectively contribute less than 40%, indicating that the marketplace's financial performance is closely tied to conditions in Southeast Brazil.

```{r}
hhi <- sum(state$revenue_share^2)
hhi
```

The Herfindahl-Hirschman Index (HHI = `r round(hhi, 3)`) reflects a moderate level of geographic concentration. While the market is not monopolistically concentrated, the heavy dependence on SP introduces regional risk — any economic disruption, logistics issue, or regulatory change in São Paulo could materially impact overall revenue.

---

## Customer Concentration

This section measures how revenue is distributed across customers and whether a small share of customers contributes a disproportionate share of revenue.

```{r}
decile_summary <- customer %>%
  group_by(decile) %>%
  summarise(
    customer_count = n(),
    total_revenue  = sum(total_revenue),
    .groups = "drop"
  ) %>%
  arrange(decile) %>%
  mutate(
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(decile_summary, aes(x = factor(decile), y = revenue_share)) +
  geom_col(fill = "#9B59B6") +
  labs(
    title = "Revenue Share by Customer Decile",
    x    = "Customer Decile (1 = Highest Spenders)",
    y    = "Revenue Share"
  ) +
  theme_minimal()
```

Revenue is heavily concentrated among the highest-value customers. The top decile alone contributes approximately 38% of total revenue, and the top three deciles together account for roughly 64%. This steep drop-off across deciles signals a high degree of customer dependency — losing a small number of high-value customers could have an outsized impact on total revenue.

```{r}
customer_sort <- customer %>%
  arrange(total_revenue) %>%
  mutate(
    cum_customer = row_number() / n(),
    cum_revenue  = cumsum(total_revenue) / sum(total_revenue)
  )
```

```{r}
ggplot(customer_sort, aes(x = cum_customer, y = cum_revenue)) +
  geom_line(color = "#E67E22", linewidth = 1.2) +
  geom_abline(linetype = "dashed") +
  labs(
    title = "Lorenz Curve of Customer Revenue Distribution",
    x = "Cumulative Share of Customers",
    y = "Cumulative Share of Revenue"
  ) +
  theme_minimal()
```

The Lorenz curve deviates substantially from the line of perfect equality, confirming a highly unequal distribution of revenue across the customer base.

```{r}
gini <- ineq(customer$total_revenue, type = "Gini")
gini
```

The Gini coefficient of `r round(gini, 3)` indicates a moderate-to-high level of customer revenue inequality. This is consistent with typical marketplace dynamics where a small cohort of repeat, high-value buyers drives the majority of revenue — but also represents a concentration risk if those customers churn.

---

## Liquidity Risk

This section evaluates whether the payment structure could create cash flow timing risk for the business.

```{r}
payment_summary <- payment %>%
  group_by(payment_type) %>%
  summarise(
    order_count   = n(),
    total_revenue = sum(order_total),
    .groups = "drop"
  ) %>%
  mutate(
    order_share   = order_count / sum(order_count),
    revenue_share = total_revenue / sum(total_revenue)
  )
```

```{r}
ggplot(payment_summary, aes(x = payment_type, y = revenue_share)) +
  geom_col(fill = "#1ABC9C") +
  labs(
    title = "Revenue Share by Payment Type",
    x = "Payment Type",
    y = "Revenue Share"
  ) +
  theme_minimal()
```

Installment payments account for a significantly larger share of revenue than one-time payments. This means cash is collected incrementally over time rather than upfront, creating a structural gap between when revenue is recognized and when it is actually received.

```{r}
weighted_installment <- sum(payment$max_installments * payment$order_total) /
  sum(payment$order_total)

weighted_installment
```

The revenue-weighted average installment count is approximately `r round(weighted_installment, 2)`, meaning that on average, each order's payment is spread across more than four installments. This installment-heavy structure implies that a meaningful portion of recognized revenue will only be collected weeks to months after the transaction — a liquidity timing risk that should be reflected in working capital planning.

---

## Final Summary

This analysis quantified four dimensions of financial risk for a Brazilian e-commerce marketplace using 100,000+ transactions from 2016–2018:

- **Revenue Stability**: Moderate volatility (CV = 0.42) with uneven short-term growth
- **Geographic Concentration**: SP drives 37% of revenue (HHI = 0.18), creating regional dependency
- **Customer Concentration**: Top 10% of customers account for 38% of revenue (Gini = 0.475)
- **Liquidity Risk**: Installment-heavy payment structure (avg. 4.12 installments) delays cash collection
- **Forecast**: Short-term revenue is expected to remain at elevated levels, with meaningful uncertainty reflecting historical volatility

Together, these findings suggest the business is in a growth phase but faces concentration risks that should be actively managed as it scales.

---

## Some Suggestions

- **CV = 0.42**: The period with the largest decline in revenue occurred during the middle of 2017 (from June to July) and from the end of 2017 to January 2018. The decline that occurred in the middle of the year might reflect a slowdown in demand after the Carnival, as Brazilian consumers' spending typically weakens after the holiday season from February to March. The decline from the end of 2017 to January 2018 coincided with the post-Black Friday and post-Christmas holiday fatigue. To stabilize revenue, I believe Olist should launch targeted promotional activities during June to July (for example, an mid-year clearance event related to the Brazilian winter season), and launch complementary discounts in January to maintain the sales momentum after the holiday period, reducing the customer churn rate from 0.42 to a more stable level of around 0.30.

- **Geographic Concentration (HHI = 0.18, SP = 37%)**: The 37% revenue share of the São Paulo Company has resulted in a significant reliance on a single region. Any local disturbances (such as logistics delays, economic downturns, or regional competition) could have a disproportionate impact on the overall revenue. MG and PR are the most feasible expansion targets. The revenue share of MG has already ranked third (approximately 12%), but it is still significantly lower than that of the São Paulo State, indicating that there are untapped demands in such an economically active state. The revenue share of PR is approximately 5%, although it is one of the wealthiest and most well-organized southern states in Brazil. Prioritizing the recruitment of sellers and conducting regional promotional activities in these two states may reduce the company's share from 37% to below 30%, thereby lowering the Herfindahl-Hirschman Index(HHI) and reducing the revenue risk in a single region.

- **Customer Concentration (Gini = 0.475)**: The Lorenz curve deviates significantly from the equality line. The top 10% of the population contributes approximately 38% of the total income. This means that losing just 1/10 of these high-value customers could immediately reduce the total income by nearly 4%. Since Olist is a market platform, repeat purchase behavior is not necessarily present, thus increasing the risk. To reduce this dependency, Olist should introduce a tiered loyalty program for the first and second tiers of customers (these two tiers account for approximately 52% of the total income): for example, providing priority customer service, exclusive seller access rights, or cash refunds for repeat purchases. The goal is to increase the repeat purchase rate of the top 20% of customers, which can both stabilize income and gradually reduce the concentration of the Gini coefficient over time.

- **Liquidity Risk (Avg installments = 4.12) **: Installment payments account for approximately 65% of the total revenue, and each order typically involves 4.12 installments. This creates a time gap between order completion and full cash settlement, limiting Olist's ability to quickly reinvest income or respond to short-term operational needs. To change the payment method, Olist should introduce a small pre-payment incentive measure. For example, one-time payment orders can enjoy a discount of 3% to 6% or free delivery benefits. Even if the proportion is slightly adjusted by 10 percentage points (from 65% of installment payments to 55%), it can significantly accelerate the cash recovery speed and shorten the average recovery period, without requiring major structural changes to the platform.
